A GPU Implementation of \gls{MVDR} for an Active Sonar System

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	14:06:07	(Mon 7.May)

The \gls{MVDR} (or Capon) beamformer is one such technique. While ensuring unit gain in the look direction, it computes the set of weights that minimizes the energy accumulated by the array from other directions. Whenever the distribution of noise and interference is primarily focused around specific angles, an array response with a high degree of suppression at these angles are chosen, resulting in an image with much improved lateral detail level.

@@

The paradigm of \gls{GPU} computing is formalized as the execution of thousand of threads, where each thread runs a copy of a common function, also know as a kernel. Threads are organized in blocks, and blocks further into a grid of blocks. The threads index in this grid usually decides the data to be process and the instruction path to go along.

A modern \gls{GPU} is a hardware architecture comprised of several hundred computing cores, each running a thread that runs a copy of a common function referred to as a kernel. This is a design that shows great promise in terms of computing power, but the tradeoff is that each computing core is left with a rather limited amount of local memory, and distributing data to the computing cores and collecting the result is a processes with some inherent latency.

data level of parallelism comes with some challenges though. First,  Data must distruted to the threads that need it, and Data must be distributed to   A fundamental limitation with such a design is that each computing core is left with a minimum of local memory, and

having access to a very limited amount of local memory an registers. This suggests that the algorithm should be decomposed into as many lightweight threads as possible. For a typical array configuration, using a single thread per pixel for \gls{MVDR} is not lightweight enough. Therefore, we seek to devise a way to decompose the methods presented in section \ref{methods}. In particular, the solutions we came up with for the different steps in the algorithm, listed in their natural order of execution, was:
\begin{enumerate}
\item \emph{Computing the spatial covariance matrix} $\eR[n]$ (equation \ref{spatialR}-\ref{finalR}). A group of threads are created that slide along the diagonals of $\eR$. In this way we can exploit the fact that entries on the diagonals overlaps across subarrays and time. 
%Finished threads can also wrap around and start on diagonals on the lower half triangular. 
In this way we process one row of $\eR$ per kernel iteration, and manage to keep the numbers of both data reads and writes at a minimum.
\item \emph{Computing} $\eRi[n]\1$ (from equation \ref{weights}). While intuition may suggest that this step is carried out by inverting $\eR$, it is better to solve the linear equation $\R\beta = \1$ for $\beta$ instead, which gives us $\beta = \eRi\1$ directly. We have tested various solvers for this task, both in-house and proprietary implementations, but achieved the best performance by using an unofficial batch solver from nVidia. Inverting $\eR$ is by far the most computationally intensive task, and a key area of focus for further improvements.
\item \emph{Computing the subarray weights} by substituting $\beta = \eRi[n]\1$ into equation \ref{weights}, $\vec w[n] = \frac{\beta}{\1\T\beta}$. A group of L threads per pixel was used to reduce shared memory pressure and to obtain coalesced reads and a minimum of writes. \emph{Finally we acquire the pixel value $z[n]$} by applying the weights to each data vector $\x_l[n]$, and sum them all up (as in equation \ref{z}). When applying weights, there are two approaches from an implementation point of view. The subarray data can, as here, be reduced to coincide in length with the subarray weights, or the weights can be extended to M in size and applied directly on $\x[n]$ and summed.  
\end{enumerate}
For the GPU, the largest bottleneck is the transfer of memory from CPU-side (Host) to the GPU-side (Device). In the upcoming results, Host to Device transfer are not taken into account. The reasoning behind this is the fact that \gls{MVDR} beamforming requires already delayed data. For efficiency, this should also be carried out on the GPU, leaving the challenging task of overcoming the limited Host-Device memory bandwidth to those who implement the delay part of delay-and-sum.

>>>>

The \gls{MVDR} (or Capon) beamformer is one such technique. While ensuring unit gain in the look direction, it computes the set of weights that minimizes the power accumulated by the array from other directions. Whenever the distribution of noise and interference is primarily focused around specific angles, an array response with a high degree of suppression at these angles are chosen, resulting in an image with much improved lateral detail level.

@@

Beamforming is a process well suited for acceleration by parallelisation, because each pixel can be processed idependently.

A modern \gls{GPU} is a hardware architecture comprised of several hundred computing cores, each running a thread that executes a copy of a common function called a kernel. However, there are some inherent drawbacks with such a design: Each core is left with a rather limited amount of local memory, and distributing data to the computing cores and collecting the result is a processes with some inherent latency.

This suggests that the algorithm should be decomposed into as many lightweight threads as possible, which are both computationally intensive and light on the memory consumption. For a typical array configuration, using a single thread per pixel is not lightweight enough. Therefore, we seek to devise a way to decompose the methods presented in section \ref{methods}. In particular, the solutions we came up with for the different steps in the algorithm, listed in their natural order of execution, was:
\begin{enumerate}
\item \emph{Computing the spatial covariance matrix} $\eR[n]$ (equation \ref{spatialR}-\ref{finalR}). A group of threads are created that slide along the diagonals of $\eR$. In this way we can exploit the fact that entries on the diagonals overlaps across subarrays and time. 
%Finished threads can also wrap around and start on diagonals on the lower half triangular. 
In this way we process one row of $\eR$ per kernel iteration, and manage to keep the numbers of both data reads and writes at a minimum.
\item \emph{Computing} $\eRi[n]\1$ (from equation \ref{weights}). While intuition may suggest that this step is carried out by inverting $\eR$, it is better to solve the linear equation $\R\beta = \1$ for $\beta$ instead, which gives us $\beta = \eRi\1$ directly. We have tested various solvers for this task, both in-house and proprietary implementations, but achieved the best performance by using an unofficial batch solver from nVidia. Inverting $\eR$ is by far the most computationally intensive task, and a key area of focus for further improvements.
\item \emph{Computing the beamformer output} $z[n]$.  by substituting $\beta = \eRi[n]\1$ into equation \ref{weights}, $\vec w[n] = \frac{\beta}{\1\T\beta}$. A group of L threads per pixel was used to reduce shared memory pressure and to obtain coalesced reads and a minimum of writes. \emph{Finally we acquire the pixel value $z[n]$} by applying the weights to each data vector $\x_l[n]$, and sum them all up (as in equation \ref{z}). When applying weights, there are two approaches from an implementation point of view. The subarray data can, as here, be reduced to coincide in length with the subarray weights, or the weights can be extended to M in size and applied directly on $\x[n]$ and summed.  
\end{enumerate}
For the GPU, the largest bottleneck is the transfer of memory from CPU-side (Host) to the GPU-side (Device). In the upcoming results, Host to Device transfer are not taken into account. The reasoning behind this is the fact that \gls{MVDR} beamforming requires already delayed data. For efficiency, this should also be carried out on the GPU, leaving the challenging task of overcoming the limited Host-Device memory bandwidth to those who implement the delay part of delay-and-sum.
